---
title: "Untitled"
output: html_document
date: "2026-05-28"
---

## Analysis dependencies

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(PathoSurveilR)
library(purrr)
library(ape)
library(tibble)
library(ggsankey)
library(ggalluvial)
library(dplyr)
library(ggplot2)
library(readr)
library(taxize)
```


## Analysis arguments

```{r}
defualt_run_output_dir <- 'output/default_runs_output'
manual_reruns_dir <- 'output/manual_runs_output'
include_perfect_sendsketch_hits <- FALSE
```


## Get list of directories with run data

These are copies of the data used as inputs for the main reports.
They contain the most important results of the pipeline and PathoSurveilR knows where to find data in these directories.

```{r}
# Get report input dirs with main results
report_input_dirs <- list.files(defualt_run_output_dir, full.names = TRUE)

# Get report input dirs for those that were manually rerun
rerun_input_dirs <- list.files(manual_reruns_dir, full.names = TRUE)

# Combine into table
input_dir_data <- tibble(
  dir_path = c(report_input_dirs, rerun_input_dirs),
  run_type = rep(c('default', 'custom'), c(length(report_input_dirs), length(rerun_input_dirs)))
)
```


## Compile results relevant to identification

This extracts statistics relevant to the interpretation of sendsketch, sourmash, and the core gene phylogeny results.
First we can define some functions that will be used by this loop.

A function to determine whether taxon names appear to be ambiguous (e.g. "unidentified bacterium"):

```{r}
is_ambiguous_taxon_name <- function(taxon_names) {
  # Get patterns to filter out
  patterns <- c(".*\\bunknown.*", ".*\\bunidentified.*", ".*incertae[_ -]+sedis.*", 
                ".*\\bambiguous\\b.*", ".*ambiguous[_ -]+taxa.*", ".*unassigned.*", 
                ".*possible.*", ".*putative.*", ".*uncultured.*", ".*\\bcandidatus.*", 
                ".*metagenome.*", ".*\\bsp\\..*", ".*\\bcf\\..*", ".*\\bendosymbiont.*" ,
                ".*\\bsymbiont.*", ".*\\bbacterium\\b.*", ".*\\bgenomosp\\..*")
  
  # Find which taxa to filter out 
  Reduce(`|`, lapply(patterns, function(x) {
    grepl(taxon_names, pattern = x, ignore.case = TRUE)
  }))
}
```

A function to return the tips in a phylogentic tree ordered by decreasing copehentic distance (how different they are based on branch length):

```{r}
get_tips_by_distance <- function(tree, target_tip) {
  # Check if target_tip exists in the tree
  if (!target_tip %in% tree$tip.label) {
    stop("Target tip not found in the tree")
  }
  
  # Get all tip labels
  all_tips <- tree$tip.label
  
  # Calculate cophenetic distances between all tips
  dist_matrix <- cophenetic.phylo(tree)
  
  # Get distances from target tip to all other tips
  distances <- dist_matrix[target_tip, ]
  
  # Sort tips by distance (excluding the target tip itself)
  sorted_tips <- names(sort(distances[names(distances) != target_tip]))
  
  # Return the sorted list
  return(sorted_tips)
}
```

Now we can loop over the input directories and add more columns of identification-relevant statistics:

```{r}
results <- pmap_dfr(input_dir_data, function(dir_path, run_type) {
  sample_metadata <- find_ps_data(dir_path, target = 'sample_metadata')[[1]]
  reference_metadata <- find_ps_data(dir_path, target = 'reference_metadata')[[1]]
  expected_species <- sample_metadata$species
  expected_genus <- strsplit(sample_metadata$species, split = ' +')[[1]][1]
  
  # Get multigene tree closest species
  if (is.null(find_ps_paths(dir_path, target = 'multigene_tree', must_exist = FALSE))) {
    multigene_tree_id <- NA_character_
  } else {
    tree <- find_ps_data(dir_path, target = 'multigene_tree')[[1]]
    ordered_tip_ids <- get_tips_by_distance(tree, sample_metadata$sample_id)
    species_id_name_key <- c(sample_metadata$species, reference_metadata$ref_name)
    names(species_id_name_key) <- c(sample_metadata$sample_id, reference_metadata$ref_id)
    ordered_tip_names <- species_id_name_key[ordered_tip_ids]
    is_ambiguous <- is_ambiguous_taxon_name(ordered_tip_names)
    multigene_tree_id <- ordered_tip_names[!is_ambiguous][1]
  }
  
  # Get sendsketch ids
  sendsketch_results <- find_ps_data(dir_path, target = 'sendsketch')[[1]]
  sendsketch_taxa <- find_ps_data(dir_path, target = 'sendsketch_taxa')[[1]]
  
  # Get sourmash ANI results
  if (is.null(find_ps_paths(dir_path, target = 'ani_matrix_csv', must_exist = FALSE))) {
    sourmash_top_id <- NA_character_
    sourmash_top_ani <- NA
    sourmash_top_correct_ref_ani <- NA
  } else {
    ani_matrix <- find_ps_data(dir_path, target = 'ani_matrix_csv')[[1]]
    sample_anis <- sort(ani_matrix[sample_metadata$sample_id, ], na.last = TRUE, decreasing = TRUE)
    sample_anis <- sample_anis[! names(sample_anis) %in% sample_metadata$sample_id]
    refs_that_match_query <- reference_metadata$ref_id[startsWith(reference_metadata$ref_name, expected_species)]
    if (length(refs_that_match_query) > 0) {
      sourmash_top_correct_ref_ani <- sample_anis[names(sample_anis) %in% refs_that_match_query][1]
    } else {
      sourmash_top_correct_ref_ani <- NA
    }
    sample_ani_names <- reference_metadata$ref_name[match(names(sample_anis), reference_metadata$ref_id)]
    not_excluded <- sample_ani_names %in% reference_metadata$ref_name[reference_metadata$ref_contextual_usage != 'excluded']
    is_ambiguous <- is_ambiguous_taxon_name(sample_ani_names)
    sample_ani_names <- sample_ani_names[not_excluded & ! is_ambiguous]
    sample_anis <- sample_anis[not_excluded]
    sourmash_top_id = sample_ani_names[1]
    sourmash_top_ani = sample_anis[1]
  }
  
  # Format row for result table
  tibble(
    sample_id = sample_metadata$sample_id,
    run_type = run_type,
    expected_species = sample_metadata$species,
    expected_genus = strsplit(expected_species, split = ' +')[[1]][1],
    expected_family = sample_metadata$family,
    expected_kingdom = sample_metadata$kingdom,
    multigene_id = multigene_tree_id,
    sourmash_id = sourmash_top_id,
    sourmash_ani = sourmash_top_ani * 100,
    sourmash_correct_ref_ani = sourmash_top_correct_ref_ani * 100,
    sendsketch_id = sendsketch_results$taxName[1],
    sendsketch_ani = as.numeric(sub(pattern = '%', replacement = '', sendsketch_results$ANI[1])),
    sendsketch_has_correct_species = any(startsWith(sendsketch_taxa$name, expected_species)),
    sendsketch_has_correct_genus = any(startsWith(sendsketch_taxa$name, expected_genus)),
    sendsketch_has_correct_family = any(startsWith(sendsketch_taxa$name, expected_family))
  )
})
```

We can now choose to filter out all analyses with perfect sendsketch hits, since these likely indicate that the assembly corresponding to the reads for the given sample is represented by its sketch in the sendsketch database.
This would interfere with our goal of simulating a newly sequenced organism.
Since sendsketch does not return the NCBI ID of the match, we can not filter for this explicitly, so this is the next best indicator.

```{r}
if (! include_perfect_sendsketch_hits)  {
  results <- results[results$sendsketch_ani < 99.9, ]  
}
```


## Interpret the identification statistics

Compare the expected taxon to the one returned by sendsketch, sourmash, and the multigene phylogeny to determine the finest taxonomic rank that was accurately classified.

```{r}
# Function to classify identification results
classify_id <- function(id, expected_species, expected_genus, expected_family, has_family_correct = FALSE) {
  if (is.na(id)) {
    return("Not run")
  }
  if (startsWith(id, expected_species)) {
    return("Species")
  }
  if (startsWith(id, expected_genus)) {
    return("Genus")
  }
  if (!is.na(has_family_correct) && has_family_correct) {
    return("Family")
  }
  return("Incorrect")
}

# Look up families for sourmash and multigene IDs using taxize
get_genus <- function(species_name) {
  if (is.na(species_name)) return(NA_character_)
  strsplit(species_name, split = ' +')[[1]][1]
}

# Get unique genera from sourmash and multigene IDs
unique_genera <- unique(na.omit(c(
  map_chr(results$sourmash_id, get_genus),
  map_chr(results$multigene_id, get_genus)
)))

# Query families using taxize
genus_families <- tax_name(sci = unique_genera, get = "family", db = "ncbi")
family_map <- setNames(genus_families$family, genus_families$query)

# Add family columns to results
results <- results %>%
  mutate(
    sourmash_family = family_map[map_chr(sourmash_id, get_genus)],
    multigene_family = family_map[map_chr(multigene_id, get_genus)]
  )

# Prepare data for Sankey plot
results <- results %>%
  mutate(
    sendsketch = pmap_chr(list(sendsketch_id, expected_species, expected_genus, expected_family, sendsketch_has_correct_family), classify_id),
    sourmash = pmap_chr(list(sourmash_id, expected_species, expected_genus, expected_family, sourmash_family == expected_family), classify_id),
    multigene = pmap_chr(list(multigene_id, expected_species, expected_genus, expected_family, multigene_family == expected_family), classify_id)
  )
```


# Add data from manual evaluation

```{r}
# Add data from human valuation
evaluation_data <- read_tsv('human_evaluation.tsv') %>%
  select(sample_id, run_type, appears_conclusive, correct_species, correct_genus, explanation)

# Only consider samples that we have results for
evaluation_data <- evaluation_data[evaluation_data$sample_id %in% results$sample_id, ]

# Add data from human evaluation onto the results table
results <- left_join(results, evaluation_data, join_by(sample_id, run_type))
results$appears_conclusive[is.na(results$appears_conclusive)] <- results$sourmash_ani[is.na(results$appears_conclusive)] >= 0.95
results$correct_species[is.na(results$correct_species)] <- results$multigene[is.na(results$correct_species)] == "Species" # Samples with the correct ID for all sendsketch, sourmash, and multigene tree were assumed correct and not manually evaluated
results$correct_genus[is.na(results$correct_genus)] <- results$multigene[is.na(results$correct_genus)] %in% c("Species", "Genus") # Samples with the correct ID for all sendsketch, sourmash, and multigene tree were assumed correct and not manually evaluated

# Function to classify human interpretation
classify_interpretation <- function(appears_conclusive, correct_species) {
  if (is.na(appears_conclusive) || is.na(correct_species)) {
    return("NA")
  }
  if (correct_species) {
    return("Correct")
  }
  if (!appears_conclusive) {
    return("Inconclusive")
  }
  return("Misleading")
}

results$interpretation = map2_chr(results$appears_conclusive, results$correct_species, classify_interpretation)
```

To manually inspect:

```{r}
results %>%
  # filter(multigene != "Species") %>%
  filter(run_type == "default") %>%
  filter(is.na(interpretation) | interpretation != "Correct" | sourmash_ani < 95) %>%
  select(sample_id, expected_species, multigene_id, sendsketch_has_correct_genus, sendsketch_has_correct_family, interpretation) %>%
  arrange(desc(interpretation))
```


## Plot how ID accuracy relates to method used

```{r}
# Get default run data
default_results <- results %>%
  filter(run_type == "default") %>%
  select(sample_id, sendsketch, sourmash, multigene, interpretation_default = interpretation, explanation_default = explanation)

# Get custom run data  
custom_results <- results %>%
  filter(run_type == "custom") %>%
  select(sample_id, interpretation_adjusted = interpretation, explanation_adjusted = explanation)

# Combine into one row per sample
sankey_data <- default_results %>%
  left_join(custom_results, by = "sample_id")

sankey_data$interpretation_adjusted[is.na(sankey_data$interpretation_adjusted)] <- sankey_data$interpretation_default[is.na(sankey_data$interpretation_adjusted)]

sankey_data$explanation <- ifelse(is.na(sankey_data$explanation_adjusted), sankey_data$explanation_default, sankey_data$explanation_adjusted)
sankey_data$explanation[is.na(sankey_data$explanation)] <- sankey_data$interpretation_adjusted[is.na(sankey_data$explanation)]
sankey_data$explanation[sankey_data$interpretation_adjusted == "Correct"] <- "Correct"

# Prepare wide-format data for ggalluvial (one row per sample)
sankey_plot_data <- sankey_data %>%
  select(sample_id, sendsketch, sourmash, multigene, interpretation_default, interpretation_adjusted, explanation)

# Color mapping
node_colors <- c(
  "Species complex" = "#7f7f7f",
  "Failed assembly" = "#7f7f7f",
  "No similar reference" = "#7f7f7f",
  "Not run" = "#7f7f7f",
  "Misleading" = "red",
  "Incorrect" = "red",
  "Family" = "orange",
  "Genus" = "#FFD700",
  "Inconclusive" = "#FFD700",
  "Species" = "#2ca02c",
  "Correct" = "#2ca02c"
)

# Convert to lodes form so each row is a stage for a sample
# to_lodes_form renumbers IDs as integers, so use an integer key for joining
sankey_plot_data$alluvium_id <- seq_len(nrow(sankey_plot_data))
lodes <- to_lodes_form(sankey_plot_data, axes = 2:7, id = "alluvium_id")
lodes$freq <- 1  # each sample contributes 1 unit

# Join interpretation_adjusted back for alluvium coloring
alluvium_colors <- sankey_plot_data %>% select(alluvium_id, interpretation_adjusted)
lodes <- lodes %>%
  left_join(alluvium_colors, by = "alluvium_id")

# Rename x positions
level_names <- c("sendsketch", "sourmash", "multigene", "interpretation_default",
                 "interpretation_adjusted", "explanation")
display_labels <- c("Sendsketch ID\nAccuracy", "Sourmash ID\nAccuracy", "Phylogeny ID\nAccuracy", "Interpretation\n(default)",
                    "Interpretation\n(adjusted)", "Explanation")
lodes$x <- factor(lodes$x, levels = level_names, labels = display_labels)
lodes$stratum <- factor(lodes$stratum, levels = rev(names(node_colors)))

# Plot with ggalluvial - each alluvium traces an individual sample
sankey_plot <- ggplot(lodes,
                       aes(x = x, y = freq, stratum = stratum, alluvium = alluvium_id)) +
  geom_alluvium(aes(fill = interpretation_adjusted), width = 1/12, alpha = 0.6, color = NA) +
  geom_stratum(aes(fill = stratum), width = 1/12, color = "black") +
  geom_text(
    stat = "stratum",
    aes(label = paste0(after_stat(stratum), " (", round(after_stat(count) / n_distinct(sankey_plot_data$sample_id) * 100, 1), "%)")),
    size = 3, color = "black", hjust = 0,
    position = position_nudge(x = 0.1)
  ) +
  scale_fill_manual(values = node_colors) +
  theme_minimal() +
  labs(title = NULL, x = NULL, y = NULL) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 12)
  )
ggsave(sankey_plot, filename = 'output/sankey_plot.svg', height = 7, width = 7)
sankey_plot
```


## Plot relationship between ANI and ID accuracy

```{r}
results %>%
  ggplot() +
  geom_point(aes(x = sourmash_ani, y = sendsketch_ani, color = interpretation))
```

