---
title: "NCBI Assembly and SRA Identification Validation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  families_per_kingdom:
    label: "Number of families to sample per kingdom"
    value: 100
  min_assemblies_per_species:
    label: "Minimum additional assemblies per species (excluding type)"
    value: 5
  max_assemblies_per_family:
    label: "Maximum assemblies to download metadata for per family"
    value: 100
  random_seed:
    label: "Random seed for reproducibility"
    value: 42
  output_path:
    label: "Output folder name for results table"
    value: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.lazy = FALSE
)

# Load required libraries
if (!require("rentrez")) install.packages("rentrez")
if (!require("dplyr")) install.packages("dplyr")
if (!require("purrr")) install.packages("purrr")
if (!require("tidyr")) install.packages("tidyr")
if (!require("jsonlite")) install.packages("jsonlite")
if (!require("httr")) install.packages("httr")
if (!require("progress")) install.packages("progress")
if (!require("tibble")) install.packages("tibble")
if (!require("readr")) install.packages("readr")

library(rentrez)
library(dplyr)
library(purrr)
library(tidyr)
library(jsonlite)
library(httr)
library(progress)
library(tibble)
library(readr)

# Create cache directory
if (!dir.exists(params$cache_dir)) {
  dir.create(params$cache_dir, recursive = TRUE)
}

# Set random seed
set.seed(params$random_seed)
```


## Randomly sample `r params$families_per_kingdom` families from each kingdom

Define the kingdom taxonomy IDs and retrieve family-level taxa for each kingdom.

```{r kingdom_taxonomy}
kingdoms <- c(
  Fungi = 4751,
  Bacteria = 2,
  Plants = 33090,
  Animals = 33208
)

family_data <- map_dfr(seq_along(kingdoms), function(i) {
  search_result <- entrez_search(
    db = "taxonomy",
    term = sprintf("txid%s[Subtree] AND family[Rank]", kingdoms[i]),
    retmax = 10000,
  )
  family_ids <- search_result$ids
  if (length(family_ids) > params$families_per_kingdom) {
    family_ids <- sample(family_ids, params$families_per_kingdom)
  }

  tibble(
    kingdom_name = names(kingdoms)[i],
    kingdom_id = kingdoms[i],
    family_id = family_ids
  )
})
```


## Step 3: Download Assembly Metadata

For each sampled family, download metadata for assemblies. Limit to `r params$max_assemblies_per_family` per family.

```{r get_assemblies}
assembly_data <- map_dfr(split(family_data, f = family_data$family_id), function(row) {
  search_result <- entrez_search(
    db = "assembly",
    term = sprintf("txid%s[Organism] AND latest[Filter]", row$family_id),
    retmax = params$max_assemblies_per_family
  )

  if (length(search_result$ids) == 0) {
    return(tibble())
  }
  
  # Fetch summary data in batches
  batch_size <- 200
  total_ids <- length(search_result$ids)
  all_summaries <- list()
  
  for (i in seq(1, total_ids, by = batch_size)) {
    end_idx <- min(i + batch_size - 1, total_ids)
    batch_ids <- search_result$ids[i:end_idx]
    
    if (i > 1) Sys.sleep(0.1)
    
    summaries <- entrez_summary(db = "assembly", id = batch_ids)
    all_summaries <- c(all_summaries, list(summaries))
  }
  
  # Extract assembly metadata
  assemblies_df <- map_dfr(all_summaries, function(summ) {
    if (is.null(summ)) return(tibble())
    
    # Handle both single and multiple summary objects
    if (inherits(summ, "esummary")) {
      summ <- list(summ)
    }
    
    map_dfr(summ, function(s) {
      tibble(
        assembly_uid = s$uid %||% NA_character_,
        assembly_accession = s$assemblyaccession %||% NA_character_,
        assembly_name = s$assemblyname %||% NA_character_,
        organism = s$organism %||% NA_character_,
        species_taxid = s$speciestaxid %||% NA_character_,
        species_name = s$speciesname %||% NA_character_,
        tax_id = s$taxid %||% NA_character_,
        assembly_type = s$assemblytype %||% NA_character_,
        biosample_id = s$biosampleid %||% NA_character_,
        coverage = s$coverage %||% NA,
        contig_n50 = s$contign50 %||% NA,
        scaffold_n50 = s$scaffoldn50 %||% NA,
        submission_date = s$submissiondate %||% NA_character_,
        gbrs_paired_asm = s$gbrspairedasm %||% NA_character_,
        refseq_category = s$refseq_category %||% NA_character_,
        assembly_status = s$assemblystatus %||% NA_character_,
        family_id = row$family_id,
        kingdom_id = row$kingdom_id,
        kingdom_name = row$kingdom_name
      )
    })
  })

  return(assemblies_df)
})
```


## Filter for species with enough assemblies and a refseq assembly

For each family, identify species that have:
2. At least `r params$min_assemblies_per_species` additional assemblies

```{r identify_candidates}
species_count <- table(assembly_data$species_taxid)
candidate_species <- names(species_count)
candidate_species <- candidate_species[species_count >= params$min_assemblies_per_species]

has_refseq <- map_lgl(candidate_species, function(species_id) {
  any(assembly_data$refseq_category[assembly_data$species_taxid == species_id] == "reference genome")
})
candidate_species <- candidate_species[has_refseq]

assembly_data <- filter(assembly_data, assembly_data$species_taxid %in% candidate_species)
```


## Get SRA accessions assocaited with biosamples

For each candidate species, verify that the type assembly has SRA data linked to the same BioSample.

```{r check_sra_links}
biosample_data <- map_dfr(assembly_data$biosample_id, function(biosample_id) {
  Sys.sleep(0.1)
  
  # Query SRA for this BioSample
  biosample_search <- entrez_search(
    db = "biosample",
    term = biosample_id,
    retmax = 1
  )
  
  # Link to SRA
  sra_links <- entrez_link(
    dbfrom = "biosample",
    db = "sra",
    id = biosample_search$ids[1]
  )
  
  if (length(sra_links$links) == 0) {
    return(tibble())
  }
  
  tibble(
    biosample_id = biosample_id, 
    sra_id = sra_links$links$biosample_sra
  )
})

# Subset to just those associated with a single SRA
sra_counts <- table(biosample_data$biosample_id) 
biosample_data <- filter(biosample_data, biosample_id %in% names(sra_counts[sra_counts == 1]))
```


## Format output for use with pathogensurveilance

```{r}
assembly_data <- left_join(assembly_data, biosample_data, by = 'biosample_id')
sample_metadata <- filter(assembly_data, ! is.na(sra_id))
sample_metadata$ncbi_accession <- sample_metadata$sra_id
sample_metadata$name <- paste('Should be', sample_metadata$species_name)
sample_metadata$description <- paste('Should be', sample_metadata$organism)
sample_metadata$ref_group_ids <- 'disallowed_refs'


reference_metadata <- tibble(
  ref_group_ids = 'disallowed_refs',
  ref_ncbi_accession = assembly_data$assembly_accession,
  ref_primary_usage = 'excluded',
  ref_contextual_usage = 'excluded'
)

if (! dir.exists(params$output_path)) {
  dir.create(params$output_path)
}
write_tsv(sample_metadata, file = file.path(params$output_path, "sample_metadata.tsv"))
write_tsv(reference_metadata, file = file.path(params$output_path, "reference_metadata.tsv"))
```


