---
title: "NCBI Assembly and SRA Identification Validation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  families_per_kingdom:
    label: "Number of families to sample per kingdom"
    value: 1000
  min_assemblies_per_species:
    label: "Minimum additional assemblies per species (excluding type)"
    value: 5
  max_assemblies_per_family:
    label: "Maximum assemblies to download metadata for per family"
    value: 10000
  random_seed:
    label: "Random seed for reproducibility"
    value: 42
  output_path:
    label: "Output folder name for results table"
    value: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  cache.lazy = FALSE
)

# Load required libraries
library(rentrez)
library(dplyr)
library(purrr)
library(tidyr)
library(jsonlite)
library(httr)
library(progress)
library(tibble)
library(readr)
library(XML)

# Set random seed
set.seed(params$random_seed)

if (! dir.exists('intermediate')) {
  dir.create('intermediate')
}
```


## Randomly sample `r params$families_per_kingdom` families from each kingdom

Define the kingdom taxonomy IDs and retrieve family-level taxa for each kingdom.

```{r kingdom_taxonomy}
kingdoms <- c(
  Fungi = 4751,
  Bacteria = 2,
  Plants = 33090,
  Animals = 33208
)

family_data_cache_path <- 'intermediate/family_data.csv' 
if (file.exists(family_data_cache_path)) {
  family_data <- read_csv(family_data_cache_path)
} else {
  family_data <- map_dfr(seq_along(kingdoms), function(i) {
    search_result <- entrez_search(
      db = "taxonomy",
      term = sprintf("txid%s[Subtree] AND family[Rank]", kingdoms[i]),
      retmax = 10000,
    )
    family_ids <- search_result$ids
    if (length(family_ids) > params$families_per_kingdom) {
      family_ids <- sample(family_ids, params$families_per_kingdom)
    }
    
    family_name <- unlist(map(split(family_ids, ceiling(seq_along(family_ids) / 50)), function(ids) {
      result <- entrez_summary('taxonomy', id = ids)
      map_chr(result, function(x) x$scientificname)
    }))
    
    tibble(
      kingdom_name = names(kingdoms)[i],
      kingdom_id = kingdoms[i],
      family_id = family_ids,
      family = family_name
    )
  })
  
  write_csv(family_data, family_data_cache_path)
}
```


## Step 3: Download Assembly Metadata

For each sampled family, download metadata for assemblies. Limit to `r params$max_assemblies_per_family` per family.

```{r get_assemblies}
all_assembly_data_cache_path <- 'intermediate/all_assembly_data.csv' 
if (file.exists(all_assembly_data_cache_path)) {
  all_assembly_data <- read_csv(all_assembly_data_cache_path)
} else {
  all_assembly_data <- map_dfr(split(family_data, f = family_data$family_id), function(row) {
    search_result <- entrez_search(
      db = "assembly",
      term = sprintf("txid%s[Organism] AND latest[Filter]", row$family_id),
      retmax = params$max_assemblies_per_family
    )
    
    if (length(search_result$ids) == 0) {
      return(tibble())
    }
    
    # Fetch summary data in batches
    batch_size <- 200
    total_ids <- length(search_result$ids)
    all_summaries <- list()
    
    for (i in seq(1, total_ids, by = batch_size)) {
      end_idx <- min(i + batch_size - 1, total_ids)
      batch_ids <- search_result$ids[i:end_idx]
      
      if (i > 1) Sys.sleep(0.1)
      
      summaries <- entrez_summary(db = "assembly", id = batch_ids)
      all_summaries <- c(all_summaries, list(summaries))
    }
    
    # Extract assembly metadata
    assemblies_df <- map_dfr(all_summaries, function(summ) {
      if (is.null(summ)) return(tibble())
      
      # Handle both single and multiple summary objects
      if (inherits(summ, "esummary")) {
        summ <- list(summ)
      }
      
      map_dfr(summ, function(s) {
        tibble(
          assembly_uid = s$uid %||% NA_character_,
          assembly_accession = s$assemblyaccession %||% NA_character_,
          assembly_name = s$assemblyname %||% NA_character_,
          organism = s$organism %||% NA_character_,
          species_id = s$speciestaxid %||% NA_character_,
          species = s$speciesname %||% NA_character_,
          tax_id = s$taxid %||% NA_character_,
          assembly_type = s$assemblytype %||% NA_character_,
          biosample_id = s$biosampleid %||% NA_character_,
          coverage = s$coverage %||% NA,
          contig_n50 = s$contign50 %||% NA,
          scaffold_n50 = s$scaffoldn50 %||% NA,
          submission_date = s$submissiondate %||% NA_character_,
          gbrs_paired_asm = s$gbrspairedasm %||% NA_character_,
          refseq_category = s$refseq_category %||% NA_character_,
          assembly_status = s$assemblystatus %||% NA_character_,
          family_id = row$family_id,
          family = row$family,
          kingdom_id = row$kingdom_id,
          kingdom = row$kingdom_name
        )
      })
    })
    
    return(assemblies_df)
  })
  
  write_csv(all_assembly_data, file = all_assembly_data_cache_path)
}
```


## Filter for species with enough assemblies and a refseq assembly

For each family, identify species that have at least `r params$min_assemblies_per_species` assemblies, including one in RefSeq.

```{r identify_candidates}
# Filter for one refseq assembly per species for species that have enough assemblies 
species_assembly_data <- map_dfr(split(all_assembly_data, all_assembly_data$species), function(species_data) {
  is_refseq <- species_data$refseq_category == "reference genome"
  if (nrow(species_data) < params$min_assemblies_per_species || all(! is_refseq)) {
    return(NULL)
  } else {
    return(species_data[which(is_refseq)[1], , drop = FALSE])
  }
})

# Filter for one assembly per genus
species_assembly_data$genus <- gsub(species_assembly_data$species_name, pattern = '([^ ]+) .*', replacement = '\\1')
species_assembly_data <- map_dfr(split(species_assembly_data, species_assembly_data$genus), function(genus_data) {
  genus_data[1, , drop = FALSE]
})
```



## Get SRA accessions assocaited with biosamples

For each candidate species, verify that the type assembly has SRA data linked to the same BioSample.

```{r check_sra_links}
# Filter for assemblies with biosample information
species_assembly_data <- filter(species_assembly_data, biosample_id != '')

# Look up SRA accessions from biosample IDs
biosample_data_cache_path <- 'intermediate/biosample_data.csv' 
if (file.exists(biosample_data_cache_path)) {
  biosample_data <- read_csv(biosample_data_cache_path)
} else {
  biosample_data <- map_dfr(species_assembly_data$biosample_id, function(biosample_id) {
    Sys.sleep(0.1)
    
    # Query SRA for this BioSample
    biosample_search <- entrez_search(
      db = "biosample",
      term = biosample_id,
      retmax = 1
    )
    Sys.sleep(0.1)
    
    # Link to SRA
    sra_links <- entrez_link(
      dbfrom = "biosample",
      db = "sra",
      id = biosample_search$ids[1]
    )
    Sys.sleep(0.1)
    
    # Subset to just those associated with a single SRA
    if (length(sra_links$links) != 1) {
      return(tibble())
    }
    
    # Get Run accessions from SRA uids
    sra_summary <- entrez_summary(db = "sra", id = sra_links$links$biosample_sra, always_return_list = TRUE)
    
    # Subset to just those associated with a single SRA run
    run_acc_count <- nchar(sra_summary[[1]]$runs) - nchar(gsub("<", "", sra_summary[[1]]$runs))
    if (length(sra_summary) != 1 || run_acc_count != 1) {
      return(tibble())
    }
    
    tibble(
      biosample_id = biosample_id, 
      sra_id = sra_links$links$biosample_sra,
      ncbi_accession = xmlAttrs(xmlRoot(xmlParse(sra_summary[[1]]$runs, asText = TRUE)))['acc']
    )
  })
  
  write_csv(biosample_data, file = biosample_data_cache_path)
}

# Subset to just those associated with a single SRA
sra_counts <- table(biosample_data$biosample_id) 
biosample_data <- filter(biosample_data, biosample_id %in% names(sra_counts[sra_counts == 1]))
```


## Format output for use with pathogensurveilance

```{r}
sample_metadata <- left_join(species_assembly_data, biosample_data, by = 'biosample_id')
sample_metadata <- filter(sample_metadata, ! is.na(ncbi_accession))
sample_metadata <- filter(sample_metadata, grepl(species, pattern = '^[a-zA-Z ]+$'))

sample_metadata$name <- paste('Should be', sample_metadata$species)
sample_metadata$description <- paste0('Should be ', sample_metadata$organism, ' (', sample_metadata$assembly_accession, ', ', sample_metadata$ncbi_accession, ')')
sample_metadata$ref_group_ids <- gsub(sample_metadata$assembly_accession, pattern = '.', replacement = '_', fixed = TRUE)
sample_metadata$ploidy <- ifelse(endsWith(sample_metadata$assembly_type, 'diploid'), 2, 1)

reference_metadata <- tibble(
  ref_id = sample_metadata$ref_group_ids,
  ref_group_ids = 'disallowed_refs',
  ref_ncbi_accession = sample_metadata$assembly_accession,
  ref_primary_usage = 'excluded',
  ref_contextual_usage = 'excluded'
)

if (! dir.exists(params$output_path)) {
  dir.create(params$output_path)
}
write_tsv(sample_metadata, file = file.path(params$output_path, "sample_metadata.tsv"))
write_tsv(reference_metadata, file = file.path(params$output_path, "reference_metadata.tsv"))
```


