---
title: "Validating pathogensurveillance reference selection"
format: html
editor: source
params:
  input: '/home/fosterz/projects/pathogensurveillance/test_profiles_out/test_bacteria'
  report_id: 'all'
  download_cache_dir: './download_cache'
  output_dir: './reference_selection_validation_results'
---

```{r}
library(RcppSimdJson)
library(purrr)
library(tibble)
library(dplyr)
library(ggplot2)
library(viridis)
library(PathoSurveilR)
```

## Required command line software

The NCBI `datasets` CLI and `sourmash` are needed to run this analysis.
The code below creates and activates a `conda` environment with this software.

```{bash}
#| eval: false
conda create -n ps_reference_selection_validation conda-forge::ncbi-datasets-cli bioconda::sourmash=4.9.4
conda activate ps_reference_selection_validation
```


##

```{r}
PathoSurveilR::print_ps_outputs(params$input)


downloaded_ref_metadata <- find_ps_data(params$input, target = 'downloaded_ref_metadata', simplify = TRUE)
sample_metadata <- find_ps_data(params$input, target = 'sample_metadata', simplify = TRUE)
ncbi_assembly_metadata <- find_ps_data(params$input, target = 'ncbi_assembly_metadata', simplify = TRUE)
```

Filter data to just the samples in this report ID. 

```{r}
ani_matrix_path_data <- find_ps_paths(params$input, target = 'ani_matrix_csv')
ani_matrix_path <- ani_matrix_path_data$path[ani_matrix_path_data$report_id == params$report_id]
ani_matrix <- find_ps_data(ani_matrix_path, target = 'ani_matrix_csv', simplify = TRUE)
sample_metadata <- sample_metadata[sample_metadata$sample_id %in% colnames(ani_matrix), , drop = FALSE]
downloaded_ref_metadata <- downloaded_ref_metadata[downloaded_ref_metadata$sample_id %in% sample_metadata$sample_id, , drop = FALSE]


```



## Prepare input data

```{bash}
#| eval: false
mkdir -p data/downloads/
mkdir -p data/assemblies/references
mkdir -p data/assemblies/samples
mkdir -p data/metadata
datasets download genome taxon 1240483 --filename data/downloads/reference_assemblies_download.zip
unzip data/downloads/reference_assemblies_download.zip -d data/downloads
mv data/downloads/ncbi_dataset/data/**/*.fna data/assemblies/references/
mv data/downloads/ncbi_dataset/data/assembly_data_report.jsonl data/metadata/sample_metadata.tsv
rm -rf data/downloads
rsync -avL fosterz@hpc.cqls.oregonstate.edu:/nfs7/BPP/Chang_Lab/Upasana/collaborations/pipeline/illumina_simulated_reads/sept17_2025_output/assemblies/spades/*scaffolds.fa.gz data/assemblies/samples/
```


## Sketch

```{bash}
mkdir -p data/signatures/references
mkdir -p data/signatures/samples
sourmash sketch dna --param-string 'scaled=1000,k=21,k=31,k=51' --outdir data/signatures/references data/assemblies/references/*
sourmash sketch dna --param-string 'scaled=1000,k=21,k=31,k=51' --outdir data/signatures/samples data/assemblies/samples/*
```


## Compare

```{bash}
sourmash compare -k 31 --ani --csv data/ani_matrix.csv --output data/ani_matrix.npy data/signatures/*/*
```

## Parse ANI matrix from Sourmash

```{r}
ani_matrix <- read.csv('data/ani_matrix.csv', header = TRUE, check.names = FALSE)
colnames(ani_matrix) <- sub(colnames(ani_matrix), pattern = '^data/assemblies/references/(.+?\\.[0-9])_.+$', replacement = '\\1')
colnames(ani_matrix) <- sub(colnames(ani_matrix), pattern = '^data/assemblies/samples/(.+?)\\.scaffolds\\.fa\\.gz$', replacement = '\\1')
rownames(ani_matrix) <- colnames(ani_matrix)
```




## Parse sample metadata

```{r}
sample_metadata <- read_tsv('data/metadata/sample_metadata.tsv')
```

## Closest refs to each sample

```{r}
n_best = 10
closet_ref_data <- map_dfr(sample_metadata$sample_id, function(sample_id) {
  ref_ani <- unlist(ani_matrix[sample_id, ! colnames(ani_matrix) %in% sample_metadata$sample_id])
  ref_ani <- ref_ani[startsWith(names(ref_ani), 'GCA_')]
  best_ref_ani <- sort(ref_ani, decreasing = TRUE)[1:n_best]
  best_refs <- paste0(names(best_ref_ani), ' (', round(best_ref_ani, 3), ')')
  mutation_rate <- as.numeric(sub(sub(sample_id, pattern = '.+?(0_?[0-9]*)$', replacement = '\\1'), pattern = '_', replacement = '.'))
  sid <- sub(sample_id, pattern = '(.+?)(0_?[0-9]*)$', replacement = '\\1')
  out <- c(sid, mutation_rate, best_refs)
  names(out) <- c('sample', 'div', '1st', '2nd', '3rd',  paste0(4:n_best, 'th'))
  out
})
write_tsv(closet_ref_data, file = 'results/closest_refs.tsv')
```

Parse selected reference data

```{r}
contextual_ref_meta_paths <- list.files(path = 'data/selections/selected', pattern = '_core_references\\.tsv$', full.names = TRUE)
mapping_ref_meta_paths <- list.files(path = 'data/selections/selected', pattern = '_mapping_references\\.tsv$', full.names = TRUE)
downloaded_ref_meta_paths <- list.files(path = 'data/selections/downloaded', full.names = TRUE)

contextual_ref_meta <- map_dfr(core_ref_meta_paths, function(path) {
  out <- suppressMessages(read_tsv(path, col_names = 'reference_id'))
  out$group_id = sub(basename(path), pattern = '^(mutation_[0-9]+)_.+$', replacement = '\\1')
  out$ref_type = 'core'
  return(out)
})

mapping_ref_meta <- map_dfr(mapping_ref_meta_paths, function(path) {
  out <- suppressMessages(read_tsv(path, col_names = c('sample_id', 'reference_id')))
  # out <- out[, 'reference_id', drop = FALSE]
  out$group_id = sub(basename(path), pattern = '^(mutation_[0-9]+)_.+$', replacement = '\\1')
  out$ref_type = 'mapping'
  return(out)
})
downloaded_ref_meta <- map_dfr(downloaded_ref_meta_paths, function(path) {
  tibble(
    reference_id = suppressMessages(read_tsv(path))$reference_id,
    sample_id = sub(basename(path), pattern = '\\.tsv$', replacement = ''),
    ref_type = 'downloaded'
  )
})
```



## Parse reference metadata

```{r}
meta_json_path <- "data/downloads/ncbi_dataset/data/assembly_data_report.jsonl"
metadata <- map_dfr(readLines(meta_json_path), function(line) {
  assem_data <- RcppSimdJson::fparse(line, always_list = TRUE)[[1]]
    attributes <- assem_data$assemblyInfo$biosample$attributes
    hosts <- paste0(attributes$value[attributes$name == 'host'], collapse = ';')
    data_parts <- list(
        reference_id = gsub(assem_data$accession, pattern = '[\\/:*?"<>| .]', replacement = '_'),
        accession = assem_data$accession,
        assembly_level = assem_data$assemblyInfo$assemblyLevel,
        assembly_status = assem_data$assemblyInfo$assemblyStatus,
        assembly_type = assem_data$assemblyInfo$assemblyType,
        hosts = ifelse(hosts == '', NA_character_, hosts),
        organism_name = gsub(assem_data$organism$organismName, pattern = '\\[|\\]', replacement = ''),
        tax_id = as.character(assem_data$organism$taxId),
        contig_l50 = as.numeric(assem_data$assemblyStats$contigL50),
        contig_n50 = as.numeric(assem_data$assemblyStats$contigN50),
        coverage = as.numeric(sub(assem_data$assemblyStats$genomeCoverage, pattern = 'x$', replacement = '')),
        number_of_component_sequences = as.numeric(assem_data$assemblyStats$numberOfComponentSequences),
        number_of_contigs = as.numeric(assem_data$assemblyStats$numberOfContigs),
        total_ungapped_length = as.numeric(assem_data$assemblyStats$totalUngappedLength),
        total_sequence_length = as.numeric(assem_data$assemblyStats$totalSequenceLength),
        source_database = assem_data$sourceDatabase,
        is_type = "typeMaterial" %in% names(assem_data),
        is_annotated = "annotationInfo" %in% names(assem_data),
        is_atypical = "atypical" %in% names(assem_data$assembly_info),
        checkm_completeness = assem_data$checkmInfo$completeness,
        checkm_contamination = assem_data$checkmInfo$contamination
    )
    data_parts[sapply(data_parts, length) == 0 | sapply(data_parts, is.null)] <- NA
    output <- as_tibble(data_parts)
})
metadata$organism_name <- gsub(metadata$organism_name, pattern = '[', replacement = '', fixed = TRUE)
metadata$organism_name <- gsub(metadata$organism_name, pattern = ']', replacement = '', fixed = TRUE)
genus_prefixes <- c('Candidatus')
genus_prefix_pattern <- paste0(genus_prefixes, collapse = '|')
binomial_pattern <- paste0('^((?:', genus_prefix_pattern, ')?) ?([a-zA-Z0-9.]+) ?([a-zA-Z0-9.]+) ?(.*)$')
metadata$species <- trimws(gsub(metadata$organism_name, pattern = binomial_pattern, replacement = '\\1 \\2 \\3', ignore.case = TRUE))
metadata$genus <- trimws(gsub(metadata$organism_name, pattern = binomial_pattern, replacement = '\\1 \\2', ignore.case = TRUE))
metadata$sample_ani <- ani_matrix[metadata$accession, 'sample']
```

## Add info on selected references 

```{r}

metadata$is_downloaded <- metadata$accession %in% downloaded_ref_ids
metadata$is_contextual <- metadata$accession %in% contextual_ref_ids
```



## Score available references on desireablity

```{r}
is_ambiguous <- function(x) {
    ambiguous_words <- c(
        'uncultured',
        'unknown',
        'incertae sedis',
        'sp\\.',
        'cf\\.',
        'endosymbiont',
        'symbiont',
        'bacterium'
    )
    ambiguous_pattern <- paste0('\\b', ambiguous_words, '\\b', collapse = '|')
    grepl(x, pattern = ambiguous_pattern, ignore.case = TRUE)
}
is_latin_binomial <- function(x) {
    grepl(x, pattern = '^[a-zA-Z]+ [a-zA-Z]+($| ).*$') & ! is_ambiguous(x)
}

metadata$is_typical <-  metadata$is_atypical == FALSE
metadata$is_refseq <- metadata$source_database == 'SOURCE_DATABASE_REFSEQ'
metadata$is_bionomial <- is_latin_binomial(metadata$species)
metadata$assembly_level <- factor(metadata$assembly_level, levels = c("Contig", "Scaffold", "Chromosome", "Complete Genome"), labels = c('Co', 'S', 'Ch', 'G'), ordered = TRUE)
metadata$purity <- 100 - metadata$checkm_contamination


priority_cols <- c(
  is_typical         = "Is Typical",
  is_type            = "Is Type Strain",
  is_refseq          = "Is RefSeq Reference",
  is_bionomial       = "Has Species Epithet",
  is_annotated       = "Is Annotated",
  assembly_level     = "Assembly Level",
  checkm_completeness = "Completeness (%)",
  purity             = "Purity (%)",
  contig_l50         = "Contig L50",
  coverage           = "Coverage (x)"
)
priority <- rev(do.call(what = order, args = metadata[names(priority_cols)]))
metadata <- metadata[priority, ]

# If both GCA and GCF are present, just show GCF
id_number_part <- sub(metadata$accession, pattern = '^GC[AF]_', replacement = '')
metadata <- metadata[! duplicated(id_number_part), ]

```




## Plot 

```{r}
plotted_cols <- setNames(c('Downloaded', 'Contextual', 'Sample ANI', priority_cols), c('is_downloaded', 'is_contextual', 'sample_ani', names(priority_cols)))
plot_data <- map_dfr(names(plotted_cols), function(col) {
  values <- metadata[[col]]
  if (is.logical(values)) {
    text <- ifelse(values, "✓", "✗")
    color_scale <- viridis::viridis(2, begin = 0.1, end = 0.9)
    color <- color_scale[as.numeric(values) + 1]
  }
  if (is.numeric(values)) {
    text <- ifelse(is.na(values), '?', format(values, digits = 3))
    color_scale <- viridis::viridis(1000, begin = 0.1, end = 0.9)
    color <- color_scale[as.integer(rescale(values, to = c(1, 1000)))]
    color[is.na(color)] <- '#999999FF'
  }
  if (is.factor(values)) {
    text <- as.character(values)
    color_scale <- viridis::viridis(4, begin = 0.1, end = 0.9)
    color <- color_scale[as.numeric(values)]
  }
  tibble(
    id = metadata$accession,
    name = metadata$organism_name,
    priority = 1:nrow(metadata),
    col = col,
    text = text,
    color = color
  )
})
plot_data$display_col <- plotted_cols[plot_data$col]
plot_data$display_col <- factor(plot_data$display_col, levels = plotted_cols, ordered = TRUE)
plot_data$display_name <- paste0(plot_data$name, ' (', plot_data$id, ')')
plot_data$display_name <- factor(plot_data$display_name, levels = rev(unique(plot_data$display_name)), ordered = TRUE)



ref_selection_plot <- plot_data %>%
  filter(priority <= 50) %>%
  ggplot(aes(x = display_col, y = display_name, fill = color, label = text)) +
  geom_tile() +
  geom_text(
    size = 3, color = "black"
  ) +
  scale_fill_manual(values = setNames(plot_data$color, plot_data$color)) + 
  scale_x_discrete(position = "top") +
  labs(x = "Selection criteria", y = "Potential reference assemblies") + 
  guides(fill = "none") +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 0, vjust = 0.5)
  )
ref_selection_plot
ggsave(ref_selection_plot, path = 'results', filename = 'plot.png', width = 8, height = 8)
```
